Verifying Fine-Grained and Lock-Free Binary Search Trees

### 2021-07-13 concurrency/Verification of Concurrent Binary Search Tree.tex
a few more edits

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in Section \ref{correctness}.

@@

\section{Correctness Proofs}
\label{correctness}
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle

@@

%look at some of the Iris papers in CPP, etc. for how we should describe the proofs

@@

Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function annotated with separation logic specification. The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body. Once we find the given key (i.e. the last $\mathsf{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agrees with the fact that the current node, represented by $\texttt{g\_current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $\texttt{g\_current}$. We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of $\mathsf{sync\_commit_same}$ lemma described in section \ref{atomicity}. This is the \emph{linearization point} of the $\mathsf{lookup}$ operation and must be done before releasing the current node's lock. In the case of spanning left or right sub-tree, we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$ and $\mathsf{else\ if}$ clause. Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\treerep(g\_root, t)$).

@@

@@

Krishna et al. \cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates. Their work is closely related to our work; we took some inspiration from them in building our ghost state. But, their proofs are on a toy language rather than real C code.
\section{Conclusion}
\label{conclusion}

>>>>

\subsection{Ghost State and Global Invariants}
We can prove memory safety and race freedom properties of shared-memory concurrent programs, by tracking the transfer of ownership of shared locations between threads that would happen if we ran the program. But, this idea is insufficient to prove that concurrently running threads are successful to accomplish some task, which is also called functional correctness. For instance, we can prove the memory safety properties of the increment example in  \ref{figure1} through the transfer of ownership of shared variable $x$ between threads. But, we can not guarantee that the value of $x$ will be $2$ after all threads complete their execution, which is demonstrated in \ref{figure1} using separation logic assertion.   
\begin{figure}[htb]
\centering
$\texttt{x = 0;}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}\qquad\pi = \pi_1\ .\ \pi_2$\\
$\begin{array}{l || l}
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{acquire(l);} & \texttt{acquire(l);}\\
\mathit{x \mapsto v \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{x++;} & \texttt{x++;}\\
\mathit{x \mapsto v+1 \land I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{x \mapsto v+1 \land I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\texttt{release(l);} & \texttt{release(l);}\\
\mathit{I_{\pi_1}(x \mapsto v \land v \geq 0)} & \mathit{I_{\pi_2}(x \mapsto v \land v \geq 0)}\\
\end{array}$\\
$\mathit{I_{\pi}(x \mapsto v \land v \geq 0)}$
\caption{The increment example annotated with separation logic assertion}
\label{figure1}
\end{figure}
We must preserve the connection between the value of shared resources and the work performed by each thread. A common approach to achieve this is to use \emph{ghost variables}, an auxiliary state introduced later in the proof to track the local information about each thread. They do not appear in the original program. Program in \ref{figure1} can be verified by creating ghost state, which tracks the latest action performed by a thread, for each thread with initial value of $0$, and update with $1$ after each thread increment the value of \texttt{x}. In VST, we can create \emph{ghost state} for any Coq type in the form of an arbitrary \emph{partial commutative monoid} (PCM), a set with a partially defined binary operation that is as associative as it can, commutative, and has a unit, as long as we can describe what happens when two elements of that type are joined together. VST provides $\mathsf {ghost\_var}$ assertion to create a simple ghost state: $\mathsf{ghost\_var}\ \mathit{sh}\ a\ g$ asserts that $g$ is a \emph{ghost name} ($\gnamety$ in Coq) associated with the value $a$, which may be of any type. We will see different kind of ghost states used to verify binary search tree in the following sections.

@@

\section{BST with Hand-over-Hand Locking}
\label{hand-over-hand}
As described in section~\ref{atomic}, we want the operations of our concurrent BSTs to satisfy atomic triples relating them to sequential operations:
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle

@@

%look at some of the Iris papers in CPP, etc. for how we should describe the proofs, e.g. section 4 of Formal Verification of a Concurrent Bounded Queue in a Weak Memory Model, Mevel and Jourdan, ICFP 2021

@@

Figure \ref{lookupproof} shows the $\mathsf{lookup}$ function annotated with separation logic specification. The proof starts with $\nodeboxrep$ and $\mathsf{atomic\_shift}$ assertions as the precondition and follows the approach similar to the $\mathsf{insert}$ proof for the verification of loop body. Once we find the given key (i.e. the last $\mathsf{else}$ clause in C code) in the tree, we need to confirm that the current state of the tree in the global invariant agrees with the fact that the current node, represented by $\texttt{g\_current}$ in the global invariant, actually contains the key-value pair $(k, v)$ in the ghost state represented $\texttt{g\_current}$. We accomplish this by opening the global invariant encoded in the $\mathsf{atomic\_shift}$, and proving that lookup satisfy the public postcondition $\mathsf{Q\_p}$ with the help of the $\mathsf{sync\_commit\_same}$ lemma described in section \ref{atomicity}. This is the \emph{linearization point} of the $\mathsf{lookup}$ operation and must be done before releasing the current node's lock. In the case of spanning left or right sub-tree, we need to establish $\mathsf{lookup\_inv}$ at the end $\mathsf{if}$ and $\mathsf{else\ if}$ clause. Here, we need to show that the key we are searching is still in the bound for left/right node, so we use $\mathsf{sync\_rollback}$ lemma to extract the bound of a left/right node encoded as the ghost state in the public precondition ($\treerep(g\_root, t)$).

@@

\section{Lock-Free BST}
We want to prove that a lock-free BST implementation satisfies the same specification as our hand-over-hand implementation. Unfortunately, provably-correct deletion in a lock-free setting is a research topic in itself (cite?), so we begin with a lock-free BST that only supports insert and lookup. Once again, we want to prove
\begin{mathpar}
\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{insert}(p, k, v)\ \langle \nodeboxrep\ p\ |\ \treerep\ (\mathrm{insert}\ t\ k\ v)\rangle

\forall t.\ \langle \nodeboxrep\ p\ |\ \treerep\ t\rangle\ \texttt{lookup}(p, k)\ \langle v.\ \nodeboxrep\ p\ |\ \treerep\ t \land \mathrm{lookup}\ t\ k = v\rangle
\end{mathpar}
though the precise definitions of $\nodeboxrep$ and $\treerep$ will differ. %something about how the fact that the specs are the same really comes across to a client -- and should we prove a client?

@@

Krishna et al. \cite{krishna2017flow} have presented the proof technique for concurrent search structure templates based on the flow framework and the Iris separation logic, and verified the implementation of concurrent B-tree, hash tables, and linked lists based on the templates. Their work is closely related to our work; we took some inspiration from them in building our ghost state. But, their proofs are on a toy language rather than real C code.

\section{Conclusion}
\label{conclusion}
